---
title: "OMIM Analysis"
author: "Molly Martorella"
date: "5/23/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, root.dir = here::here())

set.seed(1)
```

# Libraries

```{r, echo=FALSE}
library(tidyverse)
library(data.table)
library(here)
library(patchwork)
library(ComplexHeatmap)
library(ComplexUpset)
```

# Inputs

```{r}
meta_seq_path <- "data/samplekey_485.txt"
gtex_key_path <- here("data/gtex/GTEx_Analysis_2017-06-05_v8_Annotations_SampleAttributesDS.txt")

tpms_path <- "data/downsampled_counts/loseq_allpreps_threshold5000000_tpms.txt"
gtex_path <- "data/gtex/gtex_downsampled_75samps_5e+06_tpms.txt"

out_path <- here("analysis/omim/")
fig_path <- here("analysis/omim/figs/omim_lod_")

omim_path <- "data/omim/genemap2.txt"
mimtitles_path <- "data/omim/mimTitles.txt"

tis <- c("buccal", "hair", "saliva", "urine")

rep_tis_path <- "data/gtex/representative_tissues.10.txt"
```

# Functions

```{r, echo=FALSE}
source(here("src/base_functions.R"))
source(here("src/remove_txn_id.R"))

fromList <- function (input) {
  # Same as original fromList()...
  elements <- unique(unlist(input))
  data <- unlist(lapply(input, function(x) {
      x <- as.vector(match(elements, x))
      }))
  data[is.na(data)] <- as.integer(0)
  data[data != 0] <- as.integer(1)
  data <- data.frame(matrix(data, ncol = length(input), byrow = F))
  data <- data[which(rowSums(data) != 0), ]
  names(data) <- names(input)
  # ... Except now it conserves your original value names!
  row.names(data) <- elements
  return(data)
}
```

# Read in data

```{r}
meta_seq <- fread(here(meta_seq_path), data.table = FALSE)
gtex_key <- fread(gtex_key_path, data.table = FALSE)

tpms <- fread(here(tpms_path), data.table = FALSE) %>% column_to_rownames("V1")
gtex <- fread(here(gtex_path), data.table = FALSE) %>% column_to_rownames("V1")

omim <- fread(here(omim_path), data.table = FALSE) %>% janitor::clean_names()
mimtitles <- fread(here(mimtitles_path), data.table = FALSE) %>% janitor::clean_names()

gtex_tis <- fread(here(rep_tis_path), data.table = FALSE) # add kidney cortex, esophagus, spleen, lung, fibroblasts
gtex_tis <- c(gtex_tis$SMTSD, "Esophagus - Mucosa", "Kidney - Cortex", "Cells - Cultured fibroblasts", "Lung", "Spleen")

source(here::here("src/gtex_loseq_colors.R"))
myColors <- myColors[c(gtex_tis, tis)]
all_tissues <- c(gtex_tis, tis)
```

# Pre-processing

# Filter meta for samples passing cts threshold and loseq-only

```{r}
meta_seq <- meta_seq %>% filter(prep == "loseq")
tpms <- tpms[, colnames(tpms) %in% meta_seq$group_name]
meta_seq <- meta_seq %>% filter(group_name %in% colnames(tpms))

loseq_sel <- meta_seq %>% select(tissue, group_name)
```

# Format GTEx

fix txn ids, make rownames, filter for tissues, sample subset of samples, make sel, select cts samples

```{r}
# filter for samples in cts
gtex_key <- gtex_key %>% filter(SAMPID %in% colnames(gtex))

# filter for subset of tissues
gtex_sel <- gtex_key %>% 
  filter(SMTSD %in% gtex_tis) %>% dplyr::rename("tissue" = "SMTSD",
                                                "group_name" = "SAMPID") %>% 
  select(group_name, tissue)

# select cts
gtex <- gtex[, colnames(gtex) %in% gtex_sel$group_name]
```

# Merge cts/keys

```{r}
# merge cts
tpms_all <- merge(tpms, gtex, by = 0)
tpms_all <- tpms_all %>% column_to_rownames("Row.names")

# rbind keys
gtex_sel <- as.data.frame(gtex_sel)
key <- rbind(gtex_sel, loseq_sel)
```


# OMIM

# Filter for known Mendelian OMIM genes

https://www.ncbi.nlm.nih.gov/pmc/articles/PMC539987/

Asterisk --> 16388 genes (gene with known sequence) (16355 with genemap vs mim2gene)
Plus --> only 27 genes (gene with known sequence and phenotype)
Number Sign --> no genes (phenotype description, molecular basis known)
Percent --> no genes (unknown molecular basis)
NULL --> 20 genes (other, phenotypes with suspected mendelian basis) (18 with genemap)

```{r}
# nums <- mimtitles %>% filter(number_prefix == "Percent") %>% pull(mim_number)
# genes <- omim %>% filter(mim_number %in% nums) %>% filter(ensembl_gene_id != "") %>% pull(ensembl_gene_id)

gene_df <- omim %>% filter(phenotypes != "", ensembl_gene_id != "") %>% select(ensembl_gene_id, phenotypes) # 4680 genes

genes <- gene_df %>% pull(ensembl_gene_id) # 16389 genes
```

# Set up list of samples per tissue

```{r}
samps <- lapply(all_tissues, function(x) {
  key %>% filter(tissue == x) %>% pull(group_name)
})

names(samps) <- all_tissues
```

# Get gene median expression level per tissue

```{r}
meds <- data.frame(row.names = rownames(tpms_all))

for (i in 1:length(samps)) {
  
  temp <- as.matrix(tpms_all[, colnames(tpms_all) %in% samps[[i]]])
  meds[, names(samps[i])] <- matrixStats::rowMedians(temp)
  
}
```

```{r}
temp <- meds
temp[, "gene_id"] <- rownames(tpms_all)
meds_long <- temp %>% pivot_longer(cols = `Artery - Tibial`:urine, names_to = "tissue", values_to = "median_expr")
```



## UPSETR

# Plot Upset for all and select GTEx tissues

```{r}
cutoff <- 0

list_tis <- lapply(colnames(meds), function(x) {
    pass <- meds %>% select(x) %>% filter(get(x) > cutoff)
    out <- pass %>% filter(rownames(pass) %in% gene_df$ensembl_gene_id)
    rownames(out)
    })

names(list_tis) <- colnames(meds)

tis_df <- fromList(list_tis)
```

```{r}
u <- upset(tis_df, 
        c("hair", "buccal", "urine", "Whole Blood", "Kidney - Cortex", "Skin - Sun Exposed (Lower leg)"), 
        name='tissues', 
        wrap = TRUE,
        keep_empty_groups = TRUE,
        #stripes = c("white", "white", "white", "grey", "grey", "grey"),
      stripes = "white",
      min_size = 10,
      #n_intersections = 5,
      height_ratio = 0.9,
      width_ratio = 0.2,
      #group_by='sets',
      base_annotations = list('Gene Number' = intersection_size(counts = TRUE,
                                                                  text_colors = c(on_background = "black",
                                                                                  on_bar = "black"),
                                                                #text=list(vjust=-0.1, hjust=-0.1, angle=45),
                                                                  mapping=aes(fill='bars_color')) +
                                  scale_fill_manual(values = c('bars_color' = as.character(myColors)),
                                                    guide = 'none') +
                                scale_y_continuous(breaks = seq(0,3000,500))),
      matrix=(intersection_matrix(geom=geom_point(shape='circle filled', size=3))
              + scale_color_manual(values=myColors[c("hair", "buccal", "urine", "Whole Blood", "Kidney - Cortex", "Skin - Sun Exposed (Lower leg)")], guide="none")),
        queries = list(upset_query(set = "hair", fill = myColors["hair"]),
                       upset_query(set = "buccal", fill = myColors["buccal"]),
                       upset_query(set = "urine", fill = myColors["urine"]),
                       upset_query(set = "Whole Blood", fill = myColors["Whole Blood"]),
                       upset_query(set = "Kidney - Cortex", fill = myColors["Kidney - Cortex"]),
                       upset_query(set = "Skin - Sun Exposed (Lower leg)", fill = myColors["Skin - Sun Exposed (Lower leg)"])),
        set_sizes = (upset_set_size(filter_intersections=TRUE,
                                   position = 'right') + 
                       geom_text(aes(label=..count..), hjust=1.1, stat='count')) +
        theme(axis.text.x = element_text(angle=30)) +
        # geom_text(aes(label=..count..), hjust=0, stat='count', color = "white"),
        theme(plot.background = element_rect(fill='transparent', color=NA)) +
        #ggtitle(paste0(tissue)) +
        theme(plot.title = element_text(hjust = 0.5),
              axis.ticks.x = element_line(),
              axis.text.x = element_text(hjust = 1)))
```

```{r}
ggsave(u, filename = paste0(fig_path, "upset.pdf"), height = 6.5, width = 10, useDingbats = FALSE)
```



# Median expr with GTEx Heatmap

```{r}
genes_use <- intersect(rownames(meds[rowSums(meds) > 0,]), genes)
mat <- as.matrix(meds[genes_use,])

mat_scaled <- pheatmap:::scale_mat(mat, scale = "row")
mat_rm <- mat[rowSums(mat) < 1e5,] # removes hemoglobin
colnames(mat_rm) <- str_to_title(colnames(mat_rm))

myColors2 <- myColors
names(myColors2) <- str_to_title(names(myColors2))
```

```{r}
pdf(file = paste0(fig_path, "med_omim_expr_heatmap.pdf"), height = 5, width = 7)

ComplexHeatmap::Heatmap(t(mat_rm),
        #col=colorRampPalette(RColorBrewer::brewer.pal(9, "Reds"))(length(seq(0,100,1))),
        #col = colorRampPalette(RColorBrewer::brewer.pal(9, "RdBu"))(length(seq(min(mat), max(mat), 10))),
        col = circlize::colorRamp2(breaks = c(0,300), colors = c("white", "red")),
        #rect_gp = gpar(col = "white", lwd = 2),
        border = "black",
        cluster_columns = TRUE,
        cluster_rows = TRUE,
        column_title = "Median OMIM gene expression",
        show_parent_dend_line = FALSE,
        clustering_distance_rows = "euclidean",
        clustering_distance_columns = "euclidean",
        show_row_names = TRUE,
        show_column_names = FALSE,
        show_column_dend = FALSE, 
        show_row_dend = TRUE, 
        #column_dend_height = unit(4, "cm")
        row_dend_width = unit(4, "cm"),
        show_heatmap_legend = FALSE,
        row_names_gp = gpar(col = c(rep("black", 15), myColors2[c("Buccal", "Hair", "Saliva", "Urine")]), fontsize = 12),
        column_title_gp = gpar(fontsize = 12)
        )

dev.off()
```



# Iteratively filter for genes with median expression passing thresholds, intersect with OMIM vector

make df with: threshold, tissue, gene_number, total_genes, proportion

```{r}
out_df <- data.frame(row.names = all_tissues)

thresholds <- c(0.1, 1, 5, 10, 20)

for (i in tis) {
  
  temp <- meds_long %>% filter(tissue == i)
  
  for (t in thresholds) {
  
    # get rownames, intersect with omim, then take length and output to df
    
    genes_passing <- temp %>% filter(median_expr >= t) %>% pull(gene_id)
    
    out_df[i, paste0("threshold_", t)] <- length(intersect(genes_passing, genes)) 
    
  }
  
}
```

# Get genes passing threshold per sample/tissue, intersect with omim

make df with: rownames are group_name, columns are tissue, threshold_0.1 etc. with gene number filled in

```{r}
thresholds <- c(0.1, 1, 5, 10, 20)
genes_use <- intersect(genes, rownames(tpms))

test <- lapply(1:length(samps), function(x) {
  
  df <- data.frame(row.names = samps[[x]],
                   "tissue" = rep(names(samps[x]), times = length(samps[[x]])))
  
  for (s in samps[[x]]) {
    
    temp <- tpms[genes_use, s] # get omim genes and sample
    
    for (t in thresholds) {

      df[s, paste0("threshold_", t)] <- sum(temp >= t)

    }

  }
  
  return(df)
  
})

out_df_all <- rbindlist(test) # 200 samples because removes hek
```

# Format for plotting - ALL ANALYSIS

pivot longer adding threshold cagtegory column and gene_number, then summarize mean and sd gene number grouped by tissue/threshold

```{r}
plot_df_all <- out_df_all %>% pivot_longer(cols = threshold_0.1:threshold_20, names_to = "threshold", values_to = "gene_ct")

plot_df_all$total_genes <- length(genes_use)

plot_df_all$proportion <- plot_df_all$gene_ct/plot_df_all$total_genes

plot_df_all$threshold <- gsub(plot_df_all$threshold, pattern = "threshold_", replacement = "")

plot_df_all_sum <- plot_df_all %>% group_by(tissue, threshold) %>% summarise(mean_ct = round(mean(gene_ct), digits = 2),
                                                                             sd_ct = round(sd(gene_ct), digits = 2),
                                                                             mean_prop = round(mean(proportion), digits = 2),
                                                                             sd_prop = round(sd(proportion), digits = 2))
```

# Boxplot

```{r}
plot_df_all %>% 
  filter(tissue %in% tis) %>% 
  mutate(threshold = factor(threshold, levels = c("0.1", "1", "5", "10", "20"))) %>% 
  ggplot(aes(x = threshold, y = proportion, fill = tissue)) +
  geom_boxplot(outlier.shape = NA) +
  #geom_point(position = position_jitterdodge(jitter.width = 0.2), shape = 1, size = 1, alpha = 0.5) +
  theme_bw() +
  scale_fill_manual(values = myColors[tis]) +
  theme(plot.title = element_text(size = 14),
                  plot.subtitle = element_text(size = 13),
                  axis.title = element_text(size = 13),
                  axis.title.y = element_text(vjust = 2),
                  axis.title.x = element_text(vjust = 0),
                  axis.text = element_text(size = 12),
                  strip.text = element_text(size = 12),
                  legend.title = element_text(size = 13),
                  legend.text = element_text(size = 12)) +
  xlab("TPM threshold") +
  ggtitle("OMIM gene capture by noninvasive tissues") +
  scale_y_continuous(breaks = seq(0,1,0.1), limits = c(0,1))

ggsave(filename = paste0(fig_path, "all_gene_proportion.png"), height = 5, width = 6)
ggsave(filename = paste0(fig_path, "all_gene_proportion.pdf"), height = 5, width = 6, useDingbats = FALSE)
```

```{r}
plot_df_all %>% 
  filter(tissue %in% tis) %>% 
  mutate(threshold = factor(threshold, levels = c("0.1", "1", "5", "10", "20"))) %>% 
  ggplot(aes(x = threshold, y = gene_ct, fill = tissue)) +
  geom_boxplot(outlier.shape = NA) +
  #geom_point(position = position_jitterdodge(jitter.width = 0.2), shape = 1, size = 1, alpha = 0.5) +
  theme_bw() +
  scale_fill_manual(values = myColors[tis]) +
  theme(plot.title = element_text(size = 14),
                  plot.subtitle = element_text(size = 13),
                  axis.title = element_text(size = 13),
                  axis.title.y = element_text(vjust = 2),
                  axis.title.x = element_text(vjust = 0),
                  axis.text = element_text(size = 12),
                  strip.text = element_text(size = 12),
                  legend.title = element_text(size = 13),
                  legend.text = element_text(size = 12)) +
  xlab("TPM threshold") +
  ggtitle("OMIM gene capture by noninvasive tissues") +
  ylab("Gene number") +
  scale_y_continuous(breaks = seq(0, 5000, 500))

ggsave(filename = paste0(fig_path, "all_gene_number.png"), height = 5, width = 6)
ggsave(filename = paste0(fig_path, "all_gene_number.pdf"), height = 5, width = 6, useDingbats = FALSE)
```







# Format for plotting - MEDIAN ANALYSIS

```{r}
plot_df <- out_df %>% rownames_to_column("tissue") %>% pivot_longer(cols = threshold_0.1:threshold_20, names_to = "threshold", values_to = "gene_ct")

plot_df$total_genes <- length(genes_use)

plot_df$proportion <- plot_df$gene_ct/plot_df$total_genes

plot_df$threshold <- gsub(plot_df$threshold, pattern = "threshold_", replacement = "")
```

# Clustered barplot, x axis threshold, y is ct and proportion, colored by tissue

```{r}
plot_df %>% 
  filter(tissue %in% tis) %>% 
  mutate(threshold = factor(threshold, levels = c("0.1", "1", "5", "10", "20"))) %>% 
  ggplot(aes(x = threshold, y = gene_ct, fill = tissue)) +
  geom_col(position = "dodge") +
  theme_bw() + 
  scale_fill_manual(values = myColors[tis]) +
  ggtitle("OMIM genes captured") +
  ylab("Gene number") +
  xlab("Median TPM threshold") +
  scale_y_continuous(breaks = seq(0,5000,500))

ggsave(filename = paste0(fig_path, "med_gene_number.png"), width = 5, height = 4)
```

```{r}
plot_df %>% 
  filter(tissue %in% tis) %>% 
  mutate(threshold = factor(threshold, levels = c("0.1", "1", "5", "10", "20"))) %>% 
  ggplot(aes(x = threshold, y = proportion, fill = tissue)) +
  geom_col(position = "dodge") +
  theme_bw() + 
  scale_fill_manual(values = myColors[tis]) +
  ggtitle("OMIM genes captured") +
  ylab("Proportion") +
  xlab("Median TPM threshold") +
  scale_y_continuous(breaks = seq(0,1,0.1))

ggsave(filename = paste0(fig_path, "med_gene_proportion.png"), width = 5, height = 4)
```





